---
output: html_document
bibliography: FB-uncertainties.bibtex
link-citations: true
csl: nature.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
tutblog <- read.csv("birgand_tutblogs.csv", encoding = "UTF-8")
selection<-tutblog[tutblog$name=="Blog: renaming directories and files",]
```
`r paste0("# ",selection$name)`
`r paste0("### *by ",selection$authors,"*")`
`r paste0("*started ",selection$date," and updated ",format(Sys.Date(), '%Y-%m-%d'),"*")`

<br/>

### Keywords
```{r echo=FALSE, results="asis",comment=FALSE} 
cat("*",gsub(";","\n * ",selection$keywords))
```

<br/><br/>

## File database in a directory tree

I keep a database of high frequency flow and concentration data, which my colleagues and I have used to calculate uncertainties on annual nutrient and material loads [@Moatar2013-aa;@Birgand2013-dr;@Birgand2011-cp;@Birgand2011-og;@Birgand2010-qd;@Birgand2009-bw;@Moatar2009-db;@Birgand2004-fm]. Until now, I had found very practical to keep the data in *hydrological years*, which for the Northern hemisphere and in temperate climates starts around the first of september and ends late August the next year. I thus kept in the filenames, e.g., "99-00" to show that the data contained were obtained between Sep 01, 1999 and Aug 31, 2000.  

I have recently worked with daily rainfall data from Oxford, U.K., dating back from 1828. The two digit system has thus become obsolete... So I decided to change everything back to four digits for years. Two problems arose: 

1. I had lots of files to change...
2. The directory tree also contained years in the same "99-00" format just like below:

<br/>

```
[51] "/Data/Yar/MO/99-00/Lin_1min_Yar_MO_99-00.txt"                        
[52] "/Data/Yar/MO/99-00/Lin_2min_Yar_MO_99-00.txt"                        
[53] "/Data/Yar/MO/99-00/Lin_5min_Yar_MO_99-00.txt" 
```

<br/>

I decided to use the `file.rename()` function in R to deal with the problem. I found this [very good "Programatically rename files (or do other stuff to them) in R" post](https://www.r-bloggers.com/programatically-rename-files-or-do-other-stuff-to-them-in-r/), I tried to follow it but it did not work...  

<br/>

After a bit of frustration, I realized that as I was trying to rename **both the files and the directories at the same time**, and I got the angry `'No such file or directory'`. This is the reason for the post.   

<br/>

> When one needs to change the *same pattern* in the names of both directories and files, things must be done in sequence, directory names first and filenames in second.

<br/>

Now, I stumbled upon lots of mistakes and I certainly spent more time than I would have changing everything by hand... But I would not have learned and shared things.
<br/>

## Make a list of patterns

I used this code to generate a list of "00-01" and corresponding "2000-2001":

```{r pattern-list, include=TRUE}
years<-(1979:2017)
pattern<-matrix(NA,nrow = length(years)-1,2)
j = 1
for (i in head(years,-1)){
  #defining the years to be used
  pattern[j,1] = paste0(substr(as.character(i),3,4),"-",substr(as.character(i+1),3,4))
  pattern[j,2] = paste0(i,"-",i+1)
  j= j+1
}
head(pattern)
```
<br/>

## Automatically rename directories

For this, I used the code below, which contains interesting tricks
<br/>
```{r echo=TRUE, eval= FALSE}
for (i in 1:nrow(pattern)){
  is.pattern = grep(pattern[i,1],dirlist)
  if (identical(is.pattern,integer(0)) == FALSE){
    sapply(dirlist[is.pattern],FUN=function(eachPath){ 
      file.rename(from=eachPath,to= sub(pattern= pattern[i,1],replacement = pattern[i,2],eachPath))
    })
  }
}
```

<br/>

### Listing directory names
First, the code to make a list of all directories down a directory tree is, as long as the file for the code is physically located in a directory downstream of which one wants all the recursive directories. I go into more details ablout this further down.

```{r echo=TRUE, eval= FALSE}
dirlist<-list.dirs(recursive = TRUE)
```
<br/>
This yielded this:
```
[1] "."  "./A1"  "./A1/Cl"  "./A1/Cl/98-99"  "./A1/DIC" "./A1/DIC/98-99" "./A1/DOC"      
[8] "./A1/DOC/98-99" "./A1/NH4" "./A1/NH4/98-99" "./A1/NO3" "./A1/NO3/98-99" "./A1/ON"        
[15] "./A1/PO4" "./A1/PO4/98-99" "./A1/TN"  "./A1/TN/98-99"  "./A1/TP"  "./A1/TP/98-99" 
and more
```
<br/>

Then I used the `grep()` function, to list the row name of the return `dirlist`. When there was no match, R returned  `integer(0)` which means a vector of length 0. I had to only make changes when there was a match so that was in the case the return from the `grep()` function was not `integer(0)`. I found the `identical()` function that yields a boolean `TRUE` or `FALSE`.

<br/>

### Applying changes through the whole directory tree

The heart of the automatic changes is really with the `sapply()` function here, which I borrowed from the [post](https://www.r-bloggers.com/programatically-rename-files-or-do-other-stuff-to-them-in-r/). It is very smart and efficient.

<br/>
```{r echo=TRUE, eval= FALSE}
sapply(dirlist[is.pattern],FUN=function(eachPath){ 
      file.rename(from=eachPath,to= sub(pattern= pattern[i,2],replacement = pattern[i,1],eachPath))
```

<br/>
The last important thing I want to share is that to have the code work for any directory on your computer, it is better to provide absolute file names.  For this, I have chosen to add a path to files and make directory and file lists from it.  Now, there is a subtlety: Notice in the code below that I purposely did not add a `/` at the end of the path name. This is because the function `list.dirs()` automatically adds one, but not the function `list.files()`...! So to get the full file name, I just added a `/` in the code so that the file list be correct.  
<br/>

```{r echo=TRUE, eval= FALSE}
path<-"/Users/francoisbirgand/Google Drive/Echantillonnage/Methods/Data"
dirlist<-list.dirs(path,recursive = TRUE)      # recursive = TRUE means that all dir in the tree will appear

### this makes a list of all the years that need to change
years<-(1979:2017)
pattern<-matrix(NA,nrow = length(years)-1,2)
j = 1
for (i in head(years,-1)){
  #defining the years to be used
  pattern[j,1] = paste0(substr(as.character(i),3,4),"-",substr(as.character(i+1),3,4))
  pattern[j,2] = paste0(i,"-",i+1)
  j= j+1
}

### This changes all the directory names
for (i in 1:nrow(pattern)){
  is.pattern = grep(pattern[i,1],dirlist)
  if (identical(is.pattern,integer(0)) == FALSE){
    sapply(dirlist[is.pattern],FUN=function(eachPath){ 
      file.rename(from=eachPath,to= sub(pattern= pattern[i,1],replacement = pattern[i,2],eachPath))
    })
  }
}

filelist<-paste0(path,"/",list.files(path,recursive = TRUE))    # recursive = TRUE means that all files in the tree will appear

### This changes all the file names
for (i in 1:nrow(pattern)){
  is.pattern = grep(pattern[i,1],filelist)
  if (identical(is.pattern,integer(0)) == FALSE){
    sapply(filelist[is.pattern],FUN=function(eachPath){ 
      file.rename(from=eachPath,to= sub(pattern= pattern[i,1],replacement = pattern[i,2],eachPath))
    })
  }
}
```
<br/><br/><br/>

# References