---
output: html_document
bibliography: FB-uncertainties.bibtex
link-citations: true
csl: nature.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
tutblog <- read.csv("birgand_tutblogs.csv", encoding = "UTF-8")
selection<-tutblog[tutblog$name=="Blog: retrieving dates in R using strsplit",]
```
`r paste0("# ",selection$name)`
`r paste0("### *by ",selection$authors,"*")`
`r paste0("*started ",selection$date," and updated ",format(Sys.Date(), '%Y-%m-%d'),"*")`

<br/>

### Keywords
```{r echo=FALSE, results="asis",comment=FALSE} 
cat("*",gsub(";","\n * ",selection$keywords))
```

<br/><br/>

### Retrieving dates manually from many different formats

The reason for this blog is because I spent, as usual, a lot of time trying to do something very simple and I want to make sure I and others can find how I did this. The beginning is simple, I want to harmonize all my flow and water quality time series database in SI units, such that all flow rates would be in $m^3/s$ and that all concentrations would be in $mg_{elemental-unit}/L$. I have acquired all these data from our own experiments, but also from agencies and other sources. It turns out that the date formats are a total mess, and until now I corrected the format on a case by case basis, but I have had enough, it is time to have a robust format.  

<br/>
R does have functions that can read dates in, e.g. `as.Date(), strptime()`, provided that the writing format is provided and consistent. This is exactly what I do not have in my database... And I am trying to find a more universal way to read all these dates in. Below are some examples of the date format that I have had to deal with:
<br/>

```
d/m/yy h:mm
dd/mm/yy hh:mm
m/d/yyyy hh:mm:ss
etc.
```

My goal was to transform all dates into `yyyy-mm-dd hh:mm:ss`, which is a robust format that R, Excel and other programs readily understand. 
<br/><br/>

### Applying *`strsplit()`* onto a vector to extract dates and times

`strsplit()` stands for "string split" and this function will split at a given pattern, e.g., spaces, ":", "/", etc. In the example below, the result of the function is a list:
<br/>

```{r eval=TRUE,echo=TRUE}
dates<-c("8/1/91","9/1/91 1:00","10/1/91 2:30","11/1/91 4:00","12/1/91 5:30","13/1/91 7:00")
inidates=dates
datime<-strsplit(dates," ")
datime
```

<br/>

Now, I am really looking for a vector with the date and a vector with the time, and playing with lists is not that easy.  The way to do this is to use the `do.call()` function, where the list rows are bound together using `rbind()` ("row bind") just like this:
<br/><br/>

```{r eval=TRUE,echo=TRUE}
datime<-do.call(rbind,strsplit(dates," "))
datime
tempdates=datime[,1]
temptime=datime[,2]
temptime
```
<br/><br/>

Notice that in the variable `temptime`, the first value is not a time variable.  This is because in the original data, which was probably from Excel, the time values at midnight, which should be "0:00" are omitted. So the rbind function, uses what it can find to fill in a value for time when there is none. It finds a value in the first column and uses this to fill in a value in column 2. This must be taken care of later.

The next step consists in repeating the previous one with the "date column" and the "time column", once two possibilities for a general date format are provided: either "d/m/y" or "m/d/y", regardless of the nubmer of digits for each of the day, month and year.  
<br/><br/>

### Using *`formatC()`* for harmonizing all final formats

The next trick is to transform all years into "yyyy"", months into "mm", days in "dd", hours into "hh", minutes into "mm", and seconds into "ss". For this I have used the function `formatC()`.  This is typically the function that one can spend a *lot* of time looking for it... Once found, it is pretty simple. For example to transform numbers from 1 to 10 into 0001, 0002, ...., 0010, one can use this code:
<br/>

```{r echo=TRUE, eval=TRUE}
x=1:10
formatC(x,digits = 0,width = 4,format = "f",flag = "0")
```
<br/><br/>

### Final code to blindly transform dates of all formats 

The only additional tricks here is the automatic check that there is a "time" component to the initial date, that if there are no seconds, they are automatically added, and that all years are automatically transformed into "yyyy".
<br/><br/>

```{r echo=TRUE,eval=TRUE, warning=FALSE}
formatdat = "d/m/y"
      datime<-do.call( rbind , strsplit( as.character(dates) , " " ) )
      
      tempdate<-datime[,1]
      tempdate<-do.call( rbind , strsplit( as.character(tempdate) , "/" ) )

      ### Formatting years
      yr=tempdate[,3]
      yy=as.numeric(yr)
      yr[round(as.numeric(yr)/100)==1]=1900+yy[round(as.numeric(yr)/100)==1]
      yr[round(as.numeric(yr)/100)==0]=2000+yy[round(as.numeric(yr)/100)==0]

      ### Formatting days and months
      if (formatdat == "d/m/y"){
        day=formatC(as.numeric(tempdate[,1]),digits = 0,width = 2,format = "f",flag = "0")
        mon=formatC(as.numeric(tempdate[,2]),digits = 0,width = 2,format = "f",flag = "0")
      }
      if (formatdat == "m/d/y"){
        day=formatC(as.numeric(tempdate[,2]),digits = 0,width = 2,format = "f",flag = "0")
        mon=formatC(as.numeric(tempdate[,1]),digits = 0,width = 2,format = "f",flag = "0")
      }
      
      ### Formatting hours, min and sec, if they exist
      if (dim(datime)[2]==2){

        temptime<-datime[,2]
        temptime[grep("/",temptime)]="0:0"   # Replaces faulty date replacements and replaces with times
        temptime<-do.call( rbind , strsplit( as.character(temptime) , ":" ) )
        hr = formatC(as.numeric(temptime[,1]),digits = 0,width = 2,format = "f",flag = "0")
        min = formatC(as.numeric(temptime[,2]),digits = 0,width = 2,format = "f",flag = "0")
        if (dim(temptime)[2]==3){sec = formatC(as.numeric(temptime[,3]),digits = 0,width = 2,
                                               format = "f",flag = "0")}else{sec="00"}
        T= paste0(hr,":",min,":",sec)
      }else{T=rep("00:00:00",dim(datime)[1])}
      
      dates<-paste0(yr,"-",mon,"-",day," ",T)
      dates<-as.data.frame(dates)
      cbind(inidates,dates)
```