---
title: "Research Projects"
---
<br/><br/>

## Research highlights

<br/><br/>

### Biogeochemistry & Ecological Engineering

- [Water quality benefits of and the nutrient and organic matter dynamics in a constructed coastal marsh](researchNR.html) 

- Water quality benefits of a stream restoration

- Transforming Denitrifying Bioreactor Research and Applications: Unveiling the Inside of the Blackbox

<br/><br/>

### Innovations in Hydrology and Water Quality Monitoring

- High frequency water quality monitoring

- Expanding high frequency water quality to high spatial resolution

- Image-based water level monitoring: the GaugeCam project

- Image-based raingauge

<br/><br/>

### Metrology of Sampling Strategies in Watersheds

- Deriving sampling frequency charts in Brittany, France

- Deriving sampling frequency charts in North Carolina

- Evaluation of time proportional sampling for the evaluation of annual nutrient fluxes at the watershed scale

<br/><br/><br/><br/>

## Recent and current grants

<br/>

```{r echo=FALSE, results="asis",comment=FALSE, warning=FALSE, comment= FALSE, message=FALSE}
library(tidyverse)

library(lubridate)


grants <- read.csv("Birgand_grants.csv")
grants$end_date<-as.Date(grants$end_date)
recentgrants<-subset(grants, (grants$end_date-Sys.Date())>=-1000)

for (i in 1:nrow(recentgrants)){#line printing loop
  cat("* ",paste0(recentgrants[i,]$Title),". ",paste0("_",recentgrants[i,]$Funder,"_ --")," ",paste0("**$",format(sum(recentgrants[i,]$Amount), big.mark=",")),"**",paste0(" &nbsp; &nbsp; &nbsp;",month(recentgrants[i,]$start_date, label = TRUE)," ",year(recentgrants[i,]$start_date)," to ",month(recentgrants[i,]$end_date, label = TRUE)," ",year(recentgrants[i,]$end_date)),"\n", sep = "")
  cat("\n")
  
}#line print


```
<br/><br/><br/><br/>

## Past projects
<br/>

```{r echo=FALSE, results="asis",comment=FALSE, warning=FALSE, comment= FALSE, message=FALSE}

oldergrants<-subset(grants, (grants$end_date-Sys.Date())<=-1000)

for (i in 1:nrow(oldergrants)){#line printing loop
  cat("* ",paste0(oldergrants[i,]$Title),". ",paste0("_",oldergrants[i,]$Funder,"_ --")," ",paste0("**$",format(sum(oldergrants[i,]$Amount), big.mark=",")),"**",paste0(" &nbsp; &nbsp; &nbsp;",month(oldergrants[i,]$start_date, label = TRUE)," ",year(oldergrants[i,]$start_date)," to ",month(oldergrants[i,]$end_date, label = TRUE)," ",year(oldergrants[i,]$end_date)),"\n", sep = "")
  cat("\n")
  
}#line print



```